Can Auxiliary Indicators Improve COVID-19 Forecasting and Hotspot Prediction?

### 2021-06-19 forecast/paper/paper.tex
Proofed up to discussion

As described previously, the auxiliary indicators defined in terms of medical
claims (CHNG-CLI, CHNG-COVID, and DV-CLI) undergo a significant and systematic
pattern of revision, or ``backfill'', after their initial publication.  Given
their somewhat statistically-regular backfill profiles, it would be reasonable
to attempt to estimate their finalized values based on vintage data---a problem
we refer to as \textit{nowcasting}---as a pre-processing step before using them
as features in the models in \eqref{eq:forecast_ar_x}, \eqref{eq:hotspot_ar_x}. 
Nowcasting is itself a highly nontrivial modeling problem, and we do not attempt
it in this paper (it is a topic of ongoing work in our research group), but we
note that nowcasting would likely improve the performance of the models
involving claims-based signals in particular.

@@

We begin with a summary of the high-level conclusions.  Here, and in the
following subsections, we will use ``AR'' to refer to the pure autoregressive
model both in forecasting, \eqref{eq:forecast_ar}, and in hotspot prediction,
\eqref{eq:hotspot_ar} (the reference to the prediction task should always be
clear from the context). We will also use the name of an auxiliary
indicator---namely ``CHNG-CLI'',  ``CHNG-COVID'', ``CTIS-CLI-in-community'',
``DV-CLI'', or ``Google-AA''---interchangeably with the model in forecasting, 
\eqref{eq:forecast_ar_x}, or hotspot prediction, \eqref{eq:hotspot_ar_x}, that
uses this particular indicator as a feature (the meaning should be clear from
the context).  So, for example, the CHNG-CLI model in forecasting is the one in  
\eqref{eq:forecast_ar_x} that sets $X_{\ell,t}$ to be the value of the CHNG-CLI 
indicator at location $\ell$ and time $t$.
% fits the coefficients based on the training data, and makes its predictions by
% setting $X_{\ell,t}$ to the appropriate CHNG-CLI values.  
Finally, we use the term ``indicator model'' to refer to any one of the ten
models of the form \eqref{eq:forecast_ar_x}, \eqref{eq:hotspot_ar_x} (five from
each of the forecasting and hotspot prediction tasks).

@@

\item In the same aggregate view, CHNG-COVID and DV-CLI offer the biggest 
  gains in each of forecasting and hotspot prediction.  CHNG-CLI is
  inconsistent: it provides a big gain in hotspot prediction, but little gain 
  in forecasting (it seems to be hurt by a notable lack of robustness, due
  to backfill).  CTIS-CLI-in-community and Google-AA each provide decent 
  gains in forecasting and hotspot prediction.  The former's performance in 
  forecasting is notable in that it clearly improves on AR even at the largest
  ahead values.

\item In a more detailed analysis of forecasting performance, we find that the
  indicator models tend to be worse than AR when case rates are increasing (this
  is most notable in CHNG-CLI and DV-CLI), and better than AR when case rates
  are flat or decreasing (most notable in CHNG-COVID and CTIS-CLI-in-community).
  More rarely does an indicator model tend to beat AR when case rates are 
  increasing, but there appears to be some evidence of this for the Google-AA
  model. 
  % In an increasing period, when an indicator model is worse than AR, it tends
  % to underpredict relative to AR (its median forecast is lower).  In a flat
  % or decreasing period, when an indicator model is better than AR, AR tends
  % to overpredict relative to it (the median AR forecast is higher).

\item In this same analysis, when an indicator model performs better than AR in
  a decreasing period, this tends to co-occur with instances in which the
  indicator ``leads'' case rates (meaning, roughly, on a short-time scale in a
  given location, its behavior mimics that of case rates some number of days
  ahead).  On the other hand, neither leading nor lagging behavior tend to
  co-occur as consistently with an indicator model doing better in periods of
  increase, or worse in periods of increase or decrease.   
\end{itemize}

Finally, to quantify the importance of training and making predictions using
proper vintage data, we ran a parallel set of forecasting and hotspot prediction 
experiments using finalized data. The results, given in the supplement, show
that training and making predictions on finalized data can result in overly
optimistic estimates of true test-time performance (up to 10\% better in terms
of average WIS or AUC). Furthermore, since indicators can have greatly different
backfill profiles, the use of finalized data in retrospective evaluations can
even change the relative ordering of models.  For example, CHNG-CLI and DV-CLI
when trained on finalized data perform very similarly for the forecasting task.
This makes sense since they are both claims-based indicators measuring
supposedly the same thing.  However, DV-CLI outperforms CHNG-CLI on vintage
data, reflecting its has a less severe backfill profile.
% TODO note what we said here about the supplement

@@

We can also see from the figure that CHNG-COVID and DV-CLI offer the biggest
gains over AR at small ahead values, followed by CTIS-CLI-in-community and
Google-AA, with the former providing the biggest gains at large ahead values.
The CHNG-CLI model performs basically the same as AR.  This is likely due to the
fact that CHNG-CLI suffers from some sort of volatility in forecasting due to
backfill.  The evidence for this is twofold: the CHNG-CLI model benefits from a
more robust method of aggregating WIS (geometric mean; shown in the supplement),
and when we train and make predictions on all-finalized data, it provides a big
gain over AR, on par with the best-performing models (also shown in the
supplement).
% TODO note what we said here about the supplement

Figure~\ref{fig:forecast} (right panel) displays the results for hotspot
prediction, again stratified by ahead value and averaged over all HRRs and
prediction dates.  We can see many similarities to the forecasting results (now
larger AUC is better); for example, CHNG-COVID and DV-CLI offer the biggest
improvement over AR, and all models including AR degrade in performance towards
the baseline (this is now just a classifier based on random guessing, which
achieves an AUC of 0.5) as the ahead values grow, illustrating the difficulty of
the hotspot prediction problem A clear difference, however, is that the CHNG-CLI
model performs quite well in hotspot prediction, close to the best-performing
indicator models for many of the ahead values. This may be because volatility in
the CHNG-CLI indicator plays less of a role in the associated logistic model's
predicted probabilities (in general, a sigmoid function can absorb a lot of the
variability in its input).

\begin{figure*}[t]
  \includegraphics[width=\textwidth]{fig/fcast-hot-combo-1.pdf}
  \caption{Left: average WIS for each forecast model, over all forecast dates
    and all HRRs, divided by the average WIS achieved by a baseline model (a
    probabilistic version of the flat-line forecaster).  Right: area under the
    curve for each hotspot prediction model, calculated over all prediction
    dates and all HRRs.  Here and in all figures we abbreviate
    CTIS-CLI-in-community by CTIS-CLIIC.  
  }  
  \label{fig:forecast}
\end{figure*}

@@

One might ask if the benefits we observe in forecasting and hotspot
prediction have anything to do with the actual indicator themselves; could it
instead be that the indicators are simply providing \textit{implicit
  regularization} on top of  the basic AR model, when we include them as lagged
features in \eqref{eq:forecast_ar_x}, \eqref{eq:hotspot_ar_x}?

To test this hypothesis, we reran all of the prediction experiments but with
$X_{\ell,t}$ in the each indicator models in \eqref{eq:forecast_ar_x},
\eqref{eq:hotspot_ar_x} replaced with suitable random noise (bootstrap samples
from a signal's historical distribution).  The results, shown and explained more
precisely in the supplement, are vastly different (worse) than the original set
of results.  In both forecasting and hotspot prediction, the ``fake'' indicator
models (with random noise as auxiliary features) offered essentially no
improvement over the pure AR model, which provides a strong rejection
(informally speaking) of the implicit regularization hypothesis.

On the topic of regularization, it is also worth noting that the use of 
$\ell_1$ regularization (tuned by cross-validation) in fitting any of the models
in \eqref{eq:forecast_ar}, \eqref{eq:forecast_ar_x}, \eqref{eq:hotspot_ar}, 
\eqref{eq:hotspot_ar_x} did not improve their generally performance at all
(experiments not shown).  This is likely due to the fact that the number of
training samples is large compared to the number of features (6,426 training
samples and only 3--6 features).

@@

While the performance of all forecasters, including AR, will generally degrade
in up periods, different models exhibit different and interesting patterns.
CHNG-CLI, CHNG-COVID, Google-AA, and especially CTIS-CLI-in-community have
large right tails (displaying improvements over AR) over the down periods.
Google-AA and CTIS-CLI-in-community have large right tails during the flat
periods. CHNG-CLI and DV-CLI have large left tails (poor forecasts relative to
AR) in flat and up periods.  Google-AA is the only model that outperforms
the AR model, on average, over up periods.  Overall, the indicators seem to help
more during flat or down periods than up periods, with the exception of
Google-AA.

@@

The supplement pursues this analysis further.  For example, we examine
classification accuracy and log-likelihood for the hotspot task, and find a
similar phenomenon: the indicators considerably improve accuracy or
log-likelihood during flat or down periods, with more mixed behavior during up
periods, with the CHNG-CLI, CHNG-COVID, and DV-CLI leading to decreased
performance in particular.

@@

% RJT: Commented out the last paragraph for now, because this is kind of a tough
% thing to explain succintly. Can revisit if we really want to include it here,
% or just let them find it in the supplement?

@@

As described in the methods section, each of the indicators we examine could be
said to measure aspects on disease progression that would precede a positive
test. That is, we imagine that these signals should ``lead'' cases. It is
entirely reasonable to imagine that, prior to an increase of confirmed COVID-19
tests reported by a public health authority in a particular location, we would
see an increase in medical insurance claims for COVID-related outpatient
visits. However, it may well be the case that such behavior is different
during different periods. In fact, we find empirically that the ``leadingness''
of an indicator (degree to which it leads case activity) tends to be more
pronounced in down or flat periods than up periods, a plausible explanation
for the decreased performance in up periods noted above.

In the supplement, we show how to define a quantitative score to measure the
leadingness of an indicator, at any time $t$ and any location $\ell$, based on
cross correlations to case rates over a short time window around $t$.
The higher this score, the greater it ``leads'' case activity.  Figure 
\ref{fig:leading} displays correlations between the leadingness score of an 
indicator and the WIS difference (AR model minus an indicator model), stratified
by whether the target is classified as up, down, or flat.  One would expect that  
the WIS difference would be positively correlated with leadingness.  Somewhat
surprisingly, this relationship turns out to be the strongest in down periods,
and weakest in up periods.  In fact, it is very nearly the case that for each
indicator, the strength of correlations only decreases as we move from down to
flat to up  periods.  In the supplement, we extend this analysis by examining
analogous ``laggingness'' scores, but we do not find as clear patterns.

>>>>

As described previously, the auxiliary indicators defined in terms of medical
claims (CHNG-CLI, CHNG-COVID, and DV-CLI) undergo a significant and systematic
pattern of revision, or ``backfill'', after their initial publication.  Given
their somewhat statistically-regular backfill profiles, it would be reasonable
to attempt to estimate their finalized values based on vintage data---a problem
we refer to as \textit{nowcasting}---as a pre-processing step before using them
as features in the models in \eqref{eq:forecast_ar_x} and \eqref{eq:hotspot_ar_x}. 
Nowcasting is itself a highly nontrivial modeling problem, and we do not attempt
it in this paper (it is a topic of ongoing work in our research group), but we
note that nowcasting would likely improve the performance of the models
involving claims-based signals in particular.

@@

Here, and in the
following subsections, we will use ``AR'' to refer to the pure autoregressive
model both in forecasting, \eqref{eq:forecast_ar}, and in hotspot prediction,
\eqref{eq:hotspot_ar} (the reference to the prediction task should always be
clear from the context). We will also use the name of an auxiliary
indicator---namely ``CHNG-CLI'',  ``CHNG-COVID'', ``CTIS-CLI-in-community'',
``DV-CLI'', or ``Google-AA''---interchangeably with the model in forecasting, 
\eqref{eq:forecast_ar_x}, or hotspot prediction, \eqref{eq:hotspot_ar_x}, that
uses this particular indicator as a feature (the meaning should be clear from
the context).  So, for example, the CHNG-CLI model in forecasting is the one in  
\eqref{eq:forecast_ar_x} that sets $X_{\ell,t}$ to be the value of the CHNG-CLI 
indicator at location $\ell$ and time $t$.
% fits the coefficients based on the training data, and makes its predictions by
% setting $X_{\ell,t}$ to the appropriate CHNG-CLI values.  
Finally, we use the term ``indicator model'' to refer to any one of the ten
models of the form \eqref{eq:forecast_ar_x} or \eqref{eq:hotspot_ar_x} (five from
each of the forecasting and hotspot prediction tasks). We begin with a summary
of the high-level conclusions.

@@

\item In the same aggregate view, CHNG-COVID and DV-CLI offer the biggest 
  gains in both forecasting and hotspot prediction.  CHNG-CLI is
  inconsistent: it provides a big gain in hotspot prediction, but little gain 
  in forecasting (it seems to be hurt by a notable lack of robustness, due
  to backfill).  CTIS-CLI-in-community and Google-AA each provide decent 
  gains in forecasting and hotspot prediction.  The former's performance in 
  forecasting is notable in that it clearly improves on AR even at the largest
  ahead values.

\item In a more detailed analysis of forecasting performance, we find that the
  indicator models tend to be
  better than AR when case rates
  are flat or decreasing (most notable in CHNG-COVID and CTIS-CLI-in-community),
  but they are
  worse than AR when case rates are increasing (this
  is most notable in CHNG-CLI and DV-CLI).
  More rarely does an indicator model tend to beat AR when case rates are 
  increasing, but there appears to be some evidence of this for the Google-AA
  model. 
  % In an increasing period, when an indicator model is worse than AR, it tends
  % to underpredict relative to AR (its median forecast is lower).  In a flat
  % or decreasing period, when an indicator model is better than AR, AR tends
  % to overpredict relative to it (the median AR forecast is higher).

\item In this same analysis, when an indicator model performs better than AR in
  a decreasing period, this tends to co-occur with instances in which the
  indicator ``leads'' case rates (meaning, roughly, on a short-time scale in a
  given location, its behavior mimics that of future case rate behavior).  On
  the other hand, if an an indicator model does better in periods of
  increase, or worse in periods of increase or decrease, its performance is
  unrelated to leadingness.   
\end{itemize}

Finally, to quantify the importance of training and making predictions using
proper vintage data, we ran a parallel set of forecasting and hotspot prediction 
experiments using finalized data. The results, given in the supplement, show
that training and making predictions on finalized data can result in overly
optimistic estimates of true test-time performance (up to 10\% better in terms
of average WIS or AUC). Furthermore, since indicators can have greatly different
backfill profiles, the use of finalized data in retrospective evaluations
changes the relative ranking of models.  For example, CHNG-CLI and DV-CLI,
when trained on finalized data, perform very similarly for the forecasting task.
This makes sense since they are both claims-based indicators supposedly measuring
the same thing.  However, DV-CLI outperforms CHNG-CLI on vintage
data, reflecting its has a less severe backfill profile.

@@

We can also see from the figure that CHNG-COVID and DV-CLI offer the biggest
gains over AR at small ahead values, followed by CTIS-CLI-in-community and
Google-AA, with the former providing the biggest gains at large ahead values.
The CHNG-CLI model performs basically the same as AR.  This is likely due to the
fact that CHNG-CLI suffers from spatially heterogeneous volatility due to
backfill.  The evidence for this explanation is twofold: (1) the CHNG-CLI model benefits from a
more robust method of aggregating WIS (geometric mean; shown in the supplement);
and (2) when we train and make predictions on finalized data, it handily beats
AR, on par with the best-performing models (also shown in the 
supplement).
% TODO note what we said here about the supplement

Figure~\ref{fig:forecast} (right panel) displays the results for hotspot
prediction, again stratified by ahead value and averaged over all HRRs and
prediction dates.  We can see many similarities to the forecasting results (now
larger AUC is better). For example, CHNG-COVID and DV-CLI offer the biggest
improvement over AR, and all models, including AR, degrade in performance towards
the baseline (in this context, a classifier based on random guessing, which
achieves an AUC of 0.5) as the ahead values grow, illustrating the difficulty of
the hotspot prediction problem. A clear difference, however, is that the CHNG-CLI
model performs quite well in hotspot prediction, close to the best-performing
indicator models for many of the ahead values. This may be because volatility in
the CHNG-CLI indicator plays less of a role in the associated logistic model's
predicted probabilities (in general, a sigmoid function can absorb a lot of the
variability in its input).

\begin{figure*}[t]
  \includegraphics[width=\textwidth]{fig/fcast-hot-combo-1.pdf}
  \caption{Main results for both tasks. Left: average WIS for each forecast
    model, over all forecast dates 
    and all HRRs, divided by the average WIS achieved by a baseline model (a
    probabilistic version of the flat-line forecaster).  Right: area under the
    curve for each hotspot prediction model, calculated over all prediction
    dates and all HRRs.  Here and in all figures we abbreviate
    CTIS-CLI-in-community by CTIS-CLIIC.  
  }  
  \label{fig:forecast}
\end{figure*}

@@

One might ask if the benefits we observe in forecasting and hotspot
prediction have anything to do with the actual indicator themselves. A plausible
alternative explanation is that the indicators are simply providing \textit{implicit
  regularization} on top of  the basic AR model when we include them as lagged
features in \eqref{eq:forecast_ar_x} and \eqref{eq:hotspot_ar_x}.

To test this hypothesis, we reran all of the prediction experiments but with
$X_{\ell,t}$ in the each indicator models
%in \eqref{eq:forecast_ar_x} \eqref{eq:hotspot_ar_x}
replaced with suitable random noise (bootstrap samples
from a signal's historical distribution).  The results, shown and explained more
precisely in the supplement, are vastly different (worse) than the original set
of results.  In both forecasting and hotspot prediction, the ``fake'' indicator
models
% (with random noise as auxiliary features)
offered essentially no
improvement over the pure AR model, which provides a strong rejection
(informally speaking) of the implicit regularization hypothesis.

On the topic of regularization, it is also worth noting that the use of 
$\ell_1$ regularization (tuned by cross-validation) in fitting any of the models
in \eqref{eq:forecast_ar}, \eqref{eq:forecast_ar_x}, \eqref{eq:hotspot_ar}, or 
\eqref{eq:hotspot_ar_x} did not generally improve their performance
(experiments not shown).  This is likely due to the fact that the number of
training samples is large compared to the number of features (6,426 training
samples and only 3--6 features). [DJM: and that proper CV requires longer
training windows, exacerbating the nonstationarity, a scenario in which its
behavior is poorly understood.]

@@

While the performance of all forecasters, including AR, will generally degrade
in up periods, different models exhibit different and interesting patterns.
CHNG-CLI, CHNG-COVID, Google-AA, and especially CTIS-CLI-in-community have
large right tails (displaying improvements over AR) during the down periods.
Google-AA and CTIS-CLI-in-community have large right tails during the flat
periods. CHNG-CLI and DV-CLI have large left tails (poor forecasts relative to
AR) in flat and up periods.  Google-AA is the only model that outperforms
the AR model, on average, over up periods.  Overall, the indicators seem to help
more during flat or down periods than up periods, with the exception of
Google-AA.

@@

The supplement pursues this analysis further.  For example, we examine
classification accuracy and log-likelihood for the hotspot task and find a
similar phenomenon: the indicators considerably improve accuracy or
log-likelihood during flat or down periods, with more mixed behavior during up
periods when  CHNG-CLI, CHNG-COVID, and DV-CLI, in particular, lead to decreased
performance.

@@

% RJT: Commented out the last paragraph for now, because this is kind of a tough
% thing to explain succintly. Can revisit if we really want to include it here,
% or just let them find it in the supplement?
% DJM: Agreed. In the supplement.

@@

As described in the methods section, each of the indicators we examine could be
said to measure aspects of disease progression that would precede a positive
test. That is, we imagine that these signals should ``lead'' cases. It is
entirely reasonable to imagine that, prior to an increase of confirmed COVID-19
tests reported by a public health authority in a particular location, we would
see an increase in medical insurance claims for COVID-related outpatient
visits. However, it may well be the case that such behavior is different
during different periods. In fact, we find empirically that the ``leadingness''
of an indicator (degree to which it leads case activity) tends to be more
pronounced in down or flat periods than in up periods, a plausible explanation
for the decreased performance in up periods noted above.

In the supplement, we show how to define a quantitative score to measure the
leadingness of an indicator, at any time $t$ and any location $\ell$, based on
cross correlations to case rates over a short time window around $t$.
The higher this score, the greater it ``leads'' case activity.  Figure 
\ref{fig:leading} displays correlations between the leadingness score of an 
indicator and the WIS difference (AR model minus an indicator model), stratified
by whether the target is classified as up, down, or flat.  One would expect that  
the WIS difference would be positively correlated with leadingness.  Somewhat
surprisingly, this relationship turns out to be strongest in down periods
and weakest in up periods.  In fact, it is very nearly the case that for each
indicator, the strength of correlations only decreases as we move from down to
flat to up  periods.  In the supplement, we extend this analysis by examining
analogous ``laggingness'' scores, but we do not find as clear patterns.
